\textsc{RILS}-\textsc{ROLS} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-11-27 paper_resources/doc/paper.tex
Up to Section 4.4

For the (13) competitors, the results are exported from SRBench \url{https://cavalab.org/srbench/results/}, reported in~\cite{la2021contemporary}. 
The maximum computation time allowed for each run of \textsc{RILS}-\textsc{ROLS} is set to 1 hour, while the maximal number of fitness function evaluations is set to 1 million. All other algorithms from SRBench used the time limit that was equal or bellow 8 hours runtime and the same fitness evaluation limit of 1 million (see~\cite{la2021contemporary}). We did not use the full time limit of 8 hours because it was not necessary -- we empirically concluded that there were no improvements after 1 hour \textsc{RILS}-\textsc{ROLS} runtime.

@@

SRBench is an open-source benchmarking project which merge a large set of diverse benchmark datasets, contemporary SR methods as well as ML methods around a shared model evaluation and an environment for analysis, see~\cite{la2021contemporary}. It brings researchers an easy handled environment to evaluate, compare, and analyze their methods.  We involve two real-world inspired ground-truth benchmark sets that represent types of regression problems, into our computational evaluation: 
\begin{itemize}
	\item \textsc{Feynman} benchmark set, inspired from physics and formulas/models that describes   various natural laws.  
	It consists of $10^5$  samples, each described in~\cite{udrescu2020ai}. Some exact models (equations) of problem instances from this benchmark are shown in Table~\ref{tab:Feynamn-Eq}.  Among them, 116 regression problem instances are chosen in the experimental evaluation, with accordance to the results reported in~\cite{la2021contemporary}.

@@

	\item \textsc{Strogatz} benchmark set from~\cite{la2016inference}. 
	Each dataset is represented by a 2-state system of first-order, ordinary differential equations. 
	The aim of each problem is to predict rate of change of the subsequent state w.r.t\  in dependence of the current two states. These equations describes natural processes described by non-linear dynamics  exhibiting chaos.  The equations for some of the datasets that belong to this benchmark are shown in Table~\ref{table:strogatz-ODEs}. These problem instances are downloadable from \url{https://github.com/lacava/ode-strogatz}. In overall, 14 ground-truth models are used to generate the problem instances used in the comparison, also in accordance with~\cite{la2021contemporary}.

@@

\end{itemize}
The above-mentioned benchmark sets may be biased since the models which describe physical laws usually impose   symmetries, periodicity w.r.t. some variables, internal separability on some variables etc. In that way, the search process and the search space can be pre-processed and thus significantly reduced. Thereby, some methods may get significant benefit with this regard, while some others may be in a subordinate situation. 
In order to permit any kind of favoritism in the input data, we generate a set of randomly generated (non-biased) SR problem instances; this benchmark set is labelled by \textsc{Random}. These instances are generated according the following procedure. \fxnote{Opisati proceduru generisanja instanci + kombinaciju parametara za instance, itd.}

\subsection{Parameter tuning}
The \textsc{RILS}-\textsc{ROLS} algorithm employs two parameters that need to be tuned, $penalty_{size}$ and $tolerance_{error}$. The first parameter is fixed to 0.001(?), according to the preliminary experimental evaluation. The second parameter is tuned by monitoring the two errors the algorithm produced on a randomly chosen training set of xx problem instances. Among the following values \{ x, xx, ... \} we chose $tolerance_{error}=?$ for which the largest average $R^2$ and the smallest average RMSE error on all problem instances from the training set are produced. %Note that the training set is chosen randomly with a cardinalty of xx instances.

@@

\subsection{Comparison with other methods on the   ground--truth benchmark sets}
In this section, we evaluate the performance of \textsc{RILS-ROLS} algorithm and compare it to the 13 other competitors from the literature. The numerical results are reported in terms of the three Tables~\ref{tab:comp_noise0}--\ref{tab:comp_noise001}. The results for the two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Storgatz}, are given w.r.t.\   the following level of noise: 0.0 (no-noisy data), 0.001 (low-level noise), and 0.01 (high-level noise), respectively. In details, the white Gaussian noise was added to the target ($y$-axis) values as a fraction of the signal RMS value, applied also in~\cite{la2021contemporary}, i.e. for target noise level $\alpha$ we have
$$ y_{noise} = y + \epsilon, \epsilon \sim \mathcal{N}\left(0, \alpha \sqrt{\frac{1}{n} \sum {y_i^2}}\right),$$
where $n$ relates to the number of items in the input data.

Each table consists of three blocks. The first block reports the algorithm (by a row) whose performances are reported. The second block reports in two columns  percentages of the instances where the exact model has been found by respective algorithms on \textsc{Feynman} and \textsc{Strogatz} benchmark set, respectively. The third column of this block provides the overall summary exact percentage of success of the algorithm over all instances from the both benchmark sets. The third block displays the results of the respective algorithm in terms of the final $R^2$ score which indicates on accuracy of found solution -- by our criterion when the algorithm  at least reached the accuracy ($R^2$ score) of 0.999.  Note that the termination of our algorithm may not always be triggered by reaching an appropriate $R^2$ score of 0.999, but also due to the reached time limit. These results are of a particular interest as they may point out on the presence of strong over-fitting in contrast to finding the right model. This block consists of three rows that indicate the percentage of instances where respective algorithm is able to reach the desired accuracy (fulfilling the condition $R^2 \geq 0.999$ upon termination) on the both ground--truth benchmark sets separately, and the overall percentage of success rate (over all considered instances) w.r.t. the criterion of reaching the final accuracy of 0.999.

@@

In short, the exact results, thus the number of instances (in percentages) for which the output of respective algorithm is able to find the exact model, are displayed in the second block of each of the tables. The numbers of instances (in percentages) for which each algorithms deliver a precise enough model (i.e. satisfying $R^2\geq 0.999$ upon its termination) are displayed in the third block of each of the tables.

@@

\begin{itemize}
	\item Concerning the exact percentages in case no noise is utilized in the input data of the instance problems, the best performing algorithm is our \textsc{RILS-ROLS}, able to find the solution that match to the right model on 57.84\% problems instances of the benchmark set \textsc{Feynman}; the same algorithm was even more impressive in solving problem instances from benchmark set \textsc{Strogatz}, being successful on 83.57\% cases. The second best performing  approach is \textsc{AI-Feynman}, successful in solving 55.78\% problem instances from \textsc{Feynman} set, and just 27.14\%  problem instances from benchmark set   \textsc{Strogatz}, respectively. All other approaches are performing much worse, and none of them is able to solve more than 30\% instances of any of the two considered ground-truth benchmark sets. 
	\item   Concerning the exact results in presence of the   noise  level of 0.001, our \textsc{RILS-ROLS} is still performing very well, having found an exact model for 42.08\% cases considering all problem instances from the both ground-truth benchmark sets. The second best approach is \textsc{AI-Feynman} which solved in total 31.89\% problem instances; as  already noticed, this algorithm performs better when applied on the instances from \textsc{Feynman} benchmark set than it is the case of \textsc{Strogatz} benchmark set. Approach  \textsc{AFP-FE} shows to be robust w.r.t. noise so its performances only slightly deteriorate in comparison to the no-noise scenario. This puts it on the third place with overall accuracy of 21.23\%. Other approaches reach   accuracy bellow 20\%. 
	\item  Concerning the exact results when the presence of noise is large, that is the noise level of 0.01,  our \textsc{RILS-ROLS} is still performing robustly, having found an exact model for  34.77\% problem instances considering the both ground-truth benchmark sets. The second best is   \textsc{AFP-FE} reaching 20\% correct solutions. \textsc{AI-Feynman} performances are degraded significantly under this level of noise, so it is ranked only 6th overall, reaching the accuracy of 12.61\%.

	\item When comparing performances of the algorithms in terms of number of instances for which the obtained final model reaches $R^2 \geq 0.999$, in case when no noise is integrated, our \textsc{RILS}-\textsc{ROLS} is successful with the rate of success of 83.38\% (third place). However, in contrast with the exact model percentages comparisons from above, there are a few methods reaching   better percentages of success than our algorithm: \textsc{MRGP}, and \textsc{Operon}, with 92.69\%, and 86.92\%, respectively. On the other hand, the exact model percentages of these two approaches are rather low (0\% and 16\%, respectively). These two algorithms are therefore more appropriate in scenarios when the underlying (ground-truth) model is unknown, so-called black-box regression.

@@

In Figure~\ref{fig:time-avg-exact-noise}, average runtimes of the algorithms on those instance problems from the both ground-truth benchmark sets for which respective exact model has  been found, are compared. The algorithms which obtained an exact percentage of at least 5\% (considering all 1300 runs) are considered as relevant  and therefore included here, while others are omitted.  The runtimes of each algorithm  are averaged w.r.t\  different levels of noise, and presented by a bar plot. Algorithms are labeled on $x$-axis, while respective average runtimes are shown on $y$-axis. Note that $y$-axis is logarithmically scaled. The following conclusions are drawn from here.

\begin{itemize}
	\item When considering the instance problems with no noise is included, the lowest average runtime  is reported for \textsc{RILS}-\textsc{ROLS} algorithm (228$s$). A slightly worse runtime is obtained by \textsc{Dsr} algorithm (232$s$).  However, note that the exact percentage of \textsc{RILS}--\textsc{ROLS} is far better from \textsc{Dsr} (60.62\% vs. 19.71\%). 	All remaining approaches deliver an order of magnitude larger average runtimes than that of the \textsc{RILS}--\textsc{ROLS} method. 
	\item  When considering the instance problems with the noise level of 0.001,    the lowest average runtime  is again delivered by \textsc{RILS}-\textsc{ROLS} algorithm (108$s$). The second best average runtime is produced by \textsc{Dsr}, but this time significantly larger (622$s$) from the former algorithm. 
	\item When considering the instance problems with the noise level of 0.01, a similar conclusion may be drawn as that from the previous point. 
\end{itemize}

>>>>

For the (13) competitors, the results are exported from SRBench \url{https://cavalab.org/srbench/results/}, reported in~\cite{la2021contemporary}. 
The maximum computation time allowed for each run of \textsc{RILS}-\textsc{ROLS} is set to 1 hour, while the maximal number of fitness function evaluations is set to 1 million. All other algorithms from SRBench used the time limit that was equal or bellow 8 hours runtime and the same fitness evaluation limit of 1 million (see~\cite{la2021contemporary}). We did not use the full time limit of 8 hours because it was not necessary -- we empirically concluded that there almost no improvements after 1 hour \textsc{RILS}-\textsc{ROLS} runtime.

@@

SRBench is an open-source benchmarking project which merge a large set of diverse benchmark datasets, contemporary SR methods as well as ML methods around a shared model evaluation and an environment for analysis, see~\cite{la2021contemporary}. It brings researchers an easy handled environment to evaluate, compare, and analyze their methods.  We involve two real-world inspired ground-truth benchmark sets from SRBench, that represent various types of regression problems, into our computational evaluation: 
\begin{itemize}
	\item \textsc{Feynman} benchmark set, inspired from physics and formulas/models that describes   various natural laws.  
	It consists of $10^5$  samples, each described in~\cite{udrescu2020ai}. Some exact models (equations) of problem instances from this benchmark are listed in Table~\ref{tab:Feynamn-Eq}.  Among them, 116 regression problem instances are chosen in the experimental evaluation, with accordance to the results reported in~\cite{la2021contemporary} that used the same problem instances.

@@

	\item \textsc{Strogatz} benchmark set from~\cite{la2016inference}. 
	Each dataset is represented by a 2-state system of first-order, ordinary differential equations. 
	The aim of each problem is to predict rate of change of the subsequent state w.r.t\  in dependence of the current two states. These equations describes natural processes modelled by non-linear dynamics  exhibiting chaos.  The equations for some of the datasets that belong to this benchmark are shown in Table~\ref{table:strogatz-ODEs}. These problem instances are downloadable from \url{https://github.com/lacava/ode-strogatz}. In overall, 14 ground-truth models are used to generate the problem instances used in the comparison, also in accordance with~\cite{la2021contemporary}.

@@

\end{itemize}
The above-mentioned benchmark sets may be biased since the models which describe physical laws usually impose   various symmetries, periodicity w.r.t. some variables, internal separability on some variables etc. In that way, the search process and the search space can be pre-defined and thus significantly reduced. Thereby, some methods may get significant benefit with this regard, while some others may be in a subordinate situation. 
In order to permit any kind of favoritism caused by physical lows behind the input data, we generate a set of randomly generated (non-biased) SR problem instances; this benchmark set is therefore labelled by \textsc{Random}. These instances are generated according the following procedure. \fxnote{Opisati proceduru generisanja instanci + kombinaciju parametara za instance, itd.}

Another purpose of having these instances is  systematically  providing conclusions on robustness and scalability of the RILS-ROLS method w.r.t.\ various instance size parameters.

\subsection{Parameter tuning}
The \textsc{RILS}-\textsc{ROLS} algorithm employs two parameters that need to be tuned, $penalty_{size}$ and $tolerance_{error}$. The first parameter is fixed empirically to 0.001(?), according to the preliminary results.  The second parameter is tuned by monitoring the two errors the algorithm produced on a randomly chosen training set of xx problem instances. Among the following values \{ x, xx, ... \} we chose $tolerance_{error}=?$ for which the largest average $R^2$ and the smallest average RMSE error on all problem instances from the training set are produced. %Note that the training set is chosen randomly with a cardinalty of xx instances.

@@

\subsection{Comparison with other methods on the   ground--truth benchmark sets}
In this section, we evaluate our \textsc{RILS-ROLS} algorithm and compare it to the 13 other competitors from the literature. The numerical results are reported in terms of the three Tables~\ref{tab:comp_noise0}--\ref{tab:comp_noise001}. The results for the two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Storgatz}, are given w.r.t.\   the following level of noise: 0.0 (no-noisy data), 0.001 (low-level noise), and 0.01 (high-level noise), respectively. In details, the white Gaussian noise is added to the target ($y$-axis) values as a fraction of the signal RMS value, applied also in~\cite{la2021contemporary}, i.e. for target noise level $\alpha$ we have
$$ y_{noise} = y + \epsilon, \epsilon \sim \mathcal{N}\left(0, \alpha \sqrt{\frac{1}{n} \sum _{i=1}^n{y_i^2}}\right),$$
where $n$ relates to the number of items in the input data.

Each table consists of three blocks. The performances of all (14) algorithms are displayed in respective rows.  
The first block gives the label of the algorithm  whose performance is reported. The second block reports by two columns the relative percentages of all (10) algorithm runs on all problem instances for which the exact model has been found by respective algorithm  on \textsc{Feynman} and \textsc{Strogatz} benchmark set, respectively. The third column of this block provides the overall   relative exact percentage of success of the algorithm over all runs on all problem instances from the both benchmark sets. The third block displays the results of the respective algorithm in terms of   final $R^2$ score which indicate  on accuracy of found solution. The solution is considered accurate if it     produced an $R^2$ score larger that  0.999.  Note that the termination of our algorithm may not always be triggered by reaching an appropriate $R^2$ score, but also due to the reached time limit. These kind of results are of a particular interest as they may point out on the presence of strong over-fitting of algorithms in contrast to the primary goal, that is finding the right model. This block consists of three columns that provide the relative percentage of algorithm runs on the instances where he algorithm is able to reach the desired accuracy (fulfilling the condition $R^2 > 0.999$ upon termination) on the both ground--truth benchmark sets separately, and the overall percentage ov all (10) runs on all considered problem instances.   %) w.r.t. the criterion of reaching the final accuracy of 0.999.

@@

In short, the exact results, thus the number of algorithm runs over all instances (in percentages) for which the output of respective algorithm is able to find the exact model, are displayed in the second block of each  table. The numbers of runs   (in percentages) for which an algorithm  delivers a precise enough model (i.e. the outcome satisfying $R^2>  0.999$) are displayed in the third block of each of the tables.

@@

\begin{itemize}
	\item Concerning the exact percentages in case when no noise is utilized in the input data, the best performing algorithm is our \textsc{RILS-ROLS}, able to find the solution that match to the right model on 57.84\% runs over all problems instances of the benchmark set \textsc{Feynman}; it was even more impressive in solving problem instances from benchmark set \textsc{Strogatz}, being successful on 83.57\% runs. The second best performing  approach is \textsc{AI-Feynman}, successful in 55.78\%   runs over all problem instances from \textsc{Feynman} set, and just 27.14\% runs over the problem instances from benchmark set   \textsc{Strogatz}, respectively. All other approaches are performing much worse, and none of them is able to delver the exact solution for more than 30\% runs over the instances of any of the two considered ground-truth benchmark sets. 
	\item   Concerning the exact results in presence of the   noise  level of 0.001, our \textsc{RILS-ROLS} is still performing very well, having found an exact model for 42.08\% runs considering all problem instances from the both ground-truth benchmark sets. The second best approach is \textsc{AI-Feynman} which is successful on 31.89\% runs over all problem instances; as  already noticed, this algorithm performs better when applied on the instances from \textsc{Feynman} benchmark set than it is the case of \textsc{Strogatz} benchmark set. Approach  \textsc{AFP-FE} shows to be robust w.r.t. noise so its performances only slightly deteriorate in comparison to the no-noise scenario. This puts it on the third place with overall accuracy of 21.23\%. Other approaches reach   accuracy bellow 20\%. 
	\item  Concerning the exact results when the presence of noise is large with  the noise level of 0.01,  our \textsc{RILS-ROLS} is still performing robustly, having found an exact model for  34.77\% runes over all problem instances considering the both ground-truth benchmark sets. The second best is   \textsc{AFP-FE} reaching 20\% correct solutions. \textsc{AI-Feynman} performances are degraded significantly under this level of noise, so it is ranked only 6th overall, reaching the accuracy of 12.61\%.

	\item When comparing performances of the algorithms in terms of algorithm runs for which the obtained final model fulfills an accuracy of $R^2 > 0.999$, in case when no noise is integrated, our \textsc{RILS}-\textsc{ROLS} is successful with the rate of success of 83.38\% (third place). However, in contrast with the exact model percentages comparisons from above, there are a few methods reaching   better percentages of success than our algorithm: \textsc{MRGP}, and \textsc{Operon}, with 92.69\%, and 86.92\%, respectively. On the other hand, the exact model percentages of these two approaches are rather low (0\% and 16\%, respectively). These two algorithms are therefore more appropriate in scenarios when the underlying (ground-truth) model is unknown, so-called black-box regression.

@@

In Figure~\ref{fig:time-avg-exact-noise}, average runtimes of the algorithms on those instance problems from the both ground-truth benchmark sets for which respective exact model has  been found, are compared. The algorithms which obtained an exact percentage of at least 5\% (considering all 1300 runs) are considered as relevant  and therefore included here, while others are just omitted.  The runtimes of each algorithm  are averaged w.r.t\  different levels of noise, and presented by a bar plot. Algorithms are labeled on $x$-axis, while respective average runtimes (oer all runs) are shown on $y$-axis. Note that $y$-axis is logarithmically scaled. The following conclusions are drawn from here.

\begin{itemize}
	\item When considering the instance problems with no-noisy input data, the lowest average runtime  is reported for \textsc{RILS}-\textsc{ROLS} algorithm (228$s$). A slightly worse runtime is obtained by \textsc{Dsr} algorithm (232$s$).  However, note that the relative exact percentage produced by \textsc{RILS}--\textsc{ROLS} is far better from \textsc{Dsr} (60.62\% vs. 19.71\%). 	All remaining approaches deliver an order of magnitude larger average runtimes than that of the \textsc{RILS}--\textsc{ROLS} method in case of runs when the exact model has been reached.   
	\item  When considering the instance problems with the noise level of 0.001,    the lowest average runtime  to find the exact model is again delivered by \textsc{RILS}-\textsc{ROLS} algorithm (108$s$). The second best average runtime is produced by \textsc{Dsr}, but this time significantly larger (622$s$) from the former algorithm. 
	\item When considering the instance problems with the noise level of 0.01, a similar conclusion may be drawn as that from above. 
\end{itemize}
